### 2015-01-17 src/paper4/p4_methods.tex
Updated after feedback

\section{Materials and methods}

@@

We used a two-alternative forced choice (2-AFC) task to study the influence of fixation depth on the perception of linear translation. We tested two fixation depths: near (50 cm) and far (200 \si{\centi\metre}) and two different eye fixation conditions: world-, and body-fixed fixation. A trial contained two sequential motion intervals of equal duration (1s), with the motion in the same direction (either leftward or rightward). Fixation condition (world, or body-fixed) was the same in both intervals, but a different fixation depth was presented in each. Participants had to judge whether the translation during the second motion interval was longer or shorter than in the first motion interval.

@@

The timing of a single trial is shown in \figref{p4:fig1}A. Each trial started with the onset of a central fixation point (i.e. aligned between the eyes) for 0.5s. Subsequently, the first 1s motion interval commenced, during which the fixation type which was at on the the two depths, remained stationary (world) or moved along with the participant (body). Subsequently a 1.75s delay followed in which the participant was kept in complete darkness. Then, the central fixation point reappeared, followed 0.5s later by a second 1s translation interval, with the same fixation condition but with fixation point at the other depth as in the first interval. Finally, the participant responded by moving a 1-dimensional joystick away from (longer) or towards (shorter) the body. Top-view illustrations of body-centered and world-centered trials are shown in Figures 1B and 1C respectively.

@@

Across trials, we varied fixation distance, as well as the order of reference and probe, giving a total of four variations per condition (see \tabref{p4:tab1}). Movement direction (either leftward or rightward) alternated between consecutive trials. The size of the probe translation was adaptively chosen based on the participants' earlier responses (Psi method; {Kontsevich:1999tg}) to determine the point of subjective equality. This was done separately for all 16 trial types (2 main conditions x 2 reference stimuli x 2 reference/probe orders x 2 movement directions; see Table 1). A total of 25 trials were collected per trial type yielding a total of 200 trials for each of the two main conditions.

Trials were presented in two one-hour sessions. To prevent dark adaptation, we turned on the lights for 5s after each block of 8 trials, and for at least 30 s after every 2 blocks. Each of the 16 unique trial types were presented once every 2 blocks. After each block, the adaptive procedure determined which translation size to test in the following block. To increase the number of data-points available to the adaptive psychometric procedure at the beginning of the experiment, we collapsed across movement direction and reference order for the first 10 trials. After that, the procedure ran separately for each of the 16 distinct trial types.

@@

For each combination of the two main conditions and the two reference/probe orders (see \tabref{p4:tab1}), we quantified the perceived probe translation by calculating the probability the probe translaton was pereived longer than the reference translation as a function of actual probe translation. We used a maximum likelihood fit of a cumulative Gaussian function to summarize the psychometric data:

@@

in which $|x|$ represents the size of the probe translation. The mean of the Gaussian, PSE, represents the point of subjective equality. The slope of the curve reflects the precision ($1/\sigma$) of reference-probe discrimination performance. Parameter $\lambda$, representing the lapse rate, accounts for stimulus-independent errors caused by subject lapses or mistakes and was restricted to small values ($\lambda < 0.06$). Fits were performed using the Psignifit toolbox {Wichmann:2001ud, Wichmann:2001wa}.
For each trial type (see \tabref{p4:tab1}), we also quantified eye movements, corrected for drift based on initial fixation. We discarded trials containing blinks as well as trials where the final eye position was more than two standard deviations away from the condition average. In total, about 12\% of all trials were discarded. After rejecting trials, we computed the average ratio between the actual, , and ideal eye movement angle that would have occured in the world-fixed fixation condition (Equation 2). This ideal eye movement angle is computed by taking the arc-tangent of the actual translation distance, , divided by the fixation depth, . We computed this ratio, , for every condition c (see Table 1). Note that for body fixation the normalized eye movement would be  and for world fixation would be .

\begin{equation}
\label{p4:eq2}
\end{equation}

Using this ratio, , we can compute the expected eye movement,  at arbitrary translation distances, even ones we did not explicitly measure:

\begin{equation}
\label{p4:eq3}
\end{equation}

Model
Using the simple cue integration model we introduced in our previous paper (Clemens et al., 2014) as a starting point, we investigate the manner in which fixation depth is taken into account. As in our previous paper, we model the perceived distance, , as a weighted combination of a vestibular, , and oculomotor estimate of translation,  (Equation 4).

@@

As the weighting parameter, , can depend on fixation depth,, it can explain both fixation depth depedent as well as independent effects. The exact distribution between these components can be found by analyzing the weighting parameter. If the weighting parameter is the same across all values for fixation depth , then there is no depth dependent part. Similarly, highly different parameters would indicate depth dependent effect. Further note that while we assume that that the vestibular () and oculomotor weights ( sum to one, they could sum to any arbitrary value by scaling the weights appropriately. 
By definition, the probe displacement (subscript p) is perceived as equal in length to the 10 cm reference displacement (subscript r) at the PSE, . By substituting both sides by the right hand side of Equation 4, we obtain:

\begin{equation}
\label{p4:eq5}
\alpha_{d_r} \hat{\varphi}_i + (1 - \alpha_{d_r}) m_i = \alpha_{d_p} \hat{\varphi}_i + (1 - \alpha_{d_p}) m_i + \epsilon
\end{equation}

In the present experiment, the reference displacement, , was always 10 cm and the probe displacement, , was equal to the measured PSE for the presented combination of fixation types (i.e,  in Equation 1). This model (i.e. Equation 5) was then fit to data from the body and world conditions using linear regression, finding one weight for every fixation depth (that is,  and ) that minimizes the sum of squared errors (. 
As mentioned before, these weights model both depth dependent as well as depth independent effects. Perfect compensation for fixation depth would require multiplication of the eye movement angle by the actual fixation depth, . The actual value of  used depends on the relative amount of compensation for fixation depth. This means that   and, in the specific case of our near and far targets,  and . By dividing our parameters, we remove the depth independent part  and reveal the ratio of fixation depth value used for compensation, . In case of perfect compensation the expected ratio is be , while in case of no compensation it is 1.

>>>>

\section{Materials and methods}
\label{p4:sec:methods}

@@

We used a two-alternative forced choice (2-AFC) task to study the influence of fixation depth on the perception of linear translation. We tested two fixation depths: near (50 cm) and far (200 \si{\centi\metre}) and two different eye fixation conditions: world-, and body-fixed fixation. A trial contained two sequential motion intervals of equal duration (1s), with the motion in the same direction (either leftward or rightward). Fixation condition (world-, or body-fixed) was the same in both intervals, but a different fixation depth was presented in each. Participants had to judge whether the translation during the second motion interval was longer or shorter than in the first motion interval.

@@

The timing of a single trial is shown in \figref{p4:fig1}A. Each trial started with the onset of a central fixation point (i.e. aligned between the eyes) for 0.5s. Subsequently, the first 1s motion interval commenced, during which the fixation type which was at on the the two depths, remained stationary (world) or moved along with the participant (body). Subsequently a 1.75s delay followed in which the participant was kept in complete darkness. Then, the central fixation point reappeared, followed 0.5 \si{\second} later by a second 1 \si{\second} translation interval, with the same fixation condition but with the fixation point at the other depth as in the first interval. Finally, the participant responded by moving a 1-dimensional joystick away from (longer) or towards (shorter) the body. Top-view illustrations of body-centered and world-centered trials are shown in \figref{p4:fig1} panels B and C.

@@

% Should look at the 4 vs 16 in the next part, sounds confusing.
Across trials, we varied fixation distance of the reference and probe, giving a total of four variations per condition (see \tabref{p4:tab1}). Movement direction (either leftward or rightward) alternated between consecutive trials. The size of the probe translation was adaptively chosen based on the participants' earlier responses (Psi method; \citeNP{kontsevich1999}) to determine the point of subjective equality. This was done separately for all 16 trial types (2 main conditions x 2 reference stimuli x 2 reference/probe orders x 2 movement directions; see Table 1). A total of 25 trials were collected per trial type yielding a total of 200 trials for each of the two main conditions.

Trials were presented in two one-hour sessions. To prevent dark adaptation, we turned on the lights for 5 \si{\second} after each block of 8 trials, and for at least 30 s after every 2 blocks. Each of the 16 unique trial types were presented once every 2 blocks. After each block, the adaptive procedure determined which translation size to test in the following block. To increase the number of data-points available to the adaptive psychometric procedure at the beginning of the experiment, we collapsed across movement direction and reference order for the first 10 trials. After that, the procedure ran separately for each of the 16 distinct trial types.

@@

For each combination of the two main conditions and the two reference/probe orders (see \tabref{p4:tab1}), we quantified the perceived probe translation by calculating the probability the probe translaton was perceived longer than the reference translation as a function of actual probe translation. We used a maximum likelihood fit of a cumulative Gaussian function to summarize the psychometric data:

@@

in which $|x|$ represents the size of the probe translation. The mean of the Gaussian, PSE, represents the point of subjective equality. The slope of the curve reflects the precision ($1/\sigma$) of reference-probe discrimination performance. Parameter $\lambda$, representing the lapse rate, accounts for stimulus-independent errors caused by subject lapses or mistakes and was restricted to small values ($\lambda < 0.06$). Fits were performed using the Psignifit toolbox \cite{wichmann2001,wichmann2001b}.

For each trial type (see \tabref{p4:tab1}), we also quantified eye movements, corrected for drift based on initial fixation. We discarded trials containing blinks as well as trials in which final eye position exceeded two standard deviations from the condition's average. Based on these criteria, 12\% of all trials were discarded. Of the remaining trails, we computed the average ratio between the measured eye excursion, $\varphi_i$, and the angle that would be needed were the trial testing the world-fixed condition. The latter is computed by taking the arc-tangent of the actual translation distance, $m_i$, divided by the fixation depth, $d_i$, which for small $\varphi$ can be approximated by $g_c = \frac{\varphi_i m_i}{d_i}$. We computed this ratio, $g_c$, for every trial type $c$ (see \tabref{p4:tab1}). Ideally, for body-fixed trials, $g_c = 0$; and for world-fixed trials, $g_c = 1$.

\subsection{Model}

Using a straightforward model, we investigate to what extent fixation is taken into account in the contribution of eye movements to self-motion perception. As in \citeA{clemens2015a}, we model the perceived translation, $p_i$, as a weighted combination of a vestibular, $m_i$, and oculomotor estimate of translation, $\hat{\varphi}_i$ (\eqnref{p4:eq4}).

@@

% This seems a bit flaky!
Variable $i$ represents either the reference, $r$, or probe, $p$, interval. Ideally, to server a veridical self-motion cue, the brain must scale the eye movement by the depth of fixation, $d$. This scaling is reflected by parameter, $\alpha_{d_i}$. If this parameter is the same across the two fixation depths, $d$, then there is not depth modulation of the oculomotor estimate of translation.

By definition, at the PSE, the probe translation is perceived as equal in length to the 10 cm reference displacement, $p_r = p_p$. By substituting both sides by the right hand side of \eqnref{p4:eq4}, we then obtain:

\begin{equation}
\label{p4:eq5}
\alpha_{d_r} \hat{\varphi}_r + (1 - \alpha_{d_r}) m_r = \alpha_{d_p} \hat{\varphi}_p + (1 - \alpha_{d_p}) m_p + \epsilon
\end{equation}

We fit \eqnref{p4:eq5} to the data using linear regression, finding one weight for every fixation depth (that is, $\alpha_{50}$ and $\alpha_{200}$) that minimizes the sum of squared errors ($\Sigma \epsilon^2$). Because these parameters can, in theory, contain both a depth dependent and a depth independent scaling, we compute their ratio, $\frac{\alpha_{50}}{\alpha_{200}}$, to remove any depth independent components. In case of perfect compensation, the expected ratio is $\frac{50}{200} = \frac{1}{4}$, while in case of no depth-dependent scaling it is 1.

### 2015-01-17 src/paper4/p4_results.tex
Updated after feedback

We investigate the effect or near versus far fixation on eye-movement driven self-motion perception. Participants were presented with two subsequent translation movements (Figure 1) and had to judge whether the second was longer or shorter than the first. During each interval, participants fixated either a body- or world-fixed fixation point either nearby (50 cm) or far away (200 cm) from the participant.

%[While I realize that we switched the order of figure 2/3 in paper 3, I think it is better to retain the original order for paper 4].

@@

We first tested the ability of participants to follow eye movement each of the four main conditions (see \tabref{p4:tab1}). Exemplar eye traces for the 10cm reference translation each main condition are depicted in \figref{p4:fig2}A. As expected, eye movements are largely absent in the body near and body far conditions (brown and orange respectively). Following simple geometry, the eye movements were large when fixating nearby targets (purple) and small when fixating far away ones (pink). A similar pattern emerges when looking at the average normalized eye displacement (see Methods) for all participants (\figref{p4:fig2}B).  The normalized eye angle is about zero in body fixed fixations (brown and orange) and follows geometry in the world fixed fixations (purple and pink).

%[NOTE: It seems quite difficult to say something about the vergence angle; our main point concerns the different version angle between world near/far.]

@@

The performance of a one participant is illustrated in the left column of Figure 3. Each rows shows one main condition: body near versus body far (Figure 3A; brown and orange), and world near versus world far (\figref{p4:fig3}B; purple and pink). The ligher and darker colors indicate which fixation type was the reference movement (see Legend). The shift of the psychometric functions relative to the 10 \si{\centi\metre} reference (i.e. the PSE) quantifies the influence of fixation distance. For example, the rightward shift of the pink curve in Figure 3B means that for a far world fixation a longer translation (~15cm) was required for that translation to be perceived equivalent to a 10 \si{\centi\metre} reference translation with world near fixation. On the other hand, the leftward shift of the purple curve means that a shorter translation with world near fixation (~6cm) was requird for that translation to be perceived equivalent to the 10cm reference translation with world far fixation. Together, these oppositely directed shifts demonstrate that translations with world far fixation were perceived shorter than equivalent translations with world near fixation. The brown and orange curves in \figref{p4:fig3}A do not show such shift, indicating that fixation depth (i.e. near versus far) has no effect in absense of eye movements (i.e. when fixating a body fixed target).

Similar results were obtained for all participants, as shown by the individual PSEs (right column of \figref{p4:fig3}). Statistical significance of the effects of fixation depth was evaluated by comparing PSEs between the two reference conditions using a paired t-test. The PSEs were significatly different in case of world-fixed fixations, t(7) = 5..42; p < 0.01 (\figref{p4:fig3}C), no such systematic difference was found for body-fixed fixations however, t(7) = -1.17; p = 0.28 (\figref{p4:fig3}D). As in the example subject, these results indicate that fixation depth does not influence self-motion perception during body fixed fixations. Fixating further away causes self-motion to be perceived as shorted in the world-fixed fixation condition however.

@@

In order to quantify how much of the fixation depth was compensated for, we first created a simple linear model (see Methods) which attempts to explain perceived translation distance as a weighted average of a vestibular estimate (equal to the actual translation) and an oculomotor estimate (equal to the normalized eye movement times the actual translation; Equation 2).  The weighting parameter was depth dependent allowing it to compensate for fixation depth. As we used only two depths, this resulted in one weighting parameter per depth,  and  (see Table 2 for values). The predicted PSEs are plotted against the actually observed and PSEs in Figure 4. The positive correlation (ρ = 0.xx, p < 0.01) between observed and predicted PSEs shows that our simple models does reasonably well in predicting perceptual performance.

>>>>

In a recent study we have shown that eye movements serve as a cue in the perception of body translation, even in the absence of optic flow \cite{clemens2015a}. Because eye rotations must be weighed by target depth ($\varphi d = T$) to serve as an adequate translation cue, we tested the effect of near versus far fixation on self-motion perception. Participants were presented with two subsequent translation movements \figref{p4:fig1} and had to judge whether the second translation was perceived longer or shorter than the first. During each interval, participants fixated either a body- or world-fixed fixation point either nearby (50 cm) or far away (200 cm) from the participant.

@@

We first tested the ability of participants to follow the eye movement instruction in our paradigm. \figref{p4:fig2}A depicts exemplar eye traces for the 10cm reference translation with nearby and far fixation points in both the body and world condition. Changes in gaze are largely absent in the body near and body far conditions (brown and orange respectively), as required. During the world conditions, the eye movements were large when fixating nearby targets (purple) and small when fixating far away ones (pink), which reflects the geometrical constraints.

We compressed the eye movement data by taking the average ratio between the measured eye excursion and the geometrically required displacement were the target at a world-fixed location. \figref{p4:fig2}B shows these normalized eye displacements for all participants, which is about zero during body fixed fixations (brown and orange) and close to one in the world fixed fixations (purple and pink).

@@

To answer this question, \figref{p4:fig3} illustrates psychophysical data on self-motion perception of a single participant for the two-fixation depths in the body (\figref{p4:fig3}A) or world condition (\figref{p4:fig3}B) respectively. Lighter and darker colors indicate which fixation type was the reference translation (see Legend). The influence of fixation distance is characterized by a shift of the psychometric functions relative to the 10 \si{\centi\metre} reference translation (i.e. the PSE). For example, the rightward shift of the pink curve in \figref{p4:fig3}B, representing the world-condition, means that with a far target a longer translation (~15 \si{\centi\metre}) was required to be perceived equivalent to a 10 \si{\centi\metre} reference translation with nearby fixation. Likewise, the leftward shift of the purple curve indicates that a shorter translation with near fixation (~6 \si{\centi\metre}) is required to be perceived the same as the 10 \si{\centi\metre} reference translation with far fixation. Together, these opposite biases suggest that translations are perceived shorter for fixations further away. The brown and orange curves in \figref{p4:fig3}A do not show such shifts, indicating that fixation depth (i.e. near versus far) has no effect in absence of eye movements (i.e. when fixating a body-fixed target during the translation).

Similar results were obtained for all participants, as shown by the individual PSEs (right column of \figref{p4:fig3}). Statistical significance of the effects of fixation depth was evaluated by comparing PSEs for the two fixation depths using a paired t-test. PSEs were significantly different in world condition, $t(7) = 5.42$, $p < 0.01$ (\figref{p4:fig3}C), but not in the body condition, $t(7) = -1.17$, $p = 0.28$ (\figref{p4:fig3}D), confirming the single subject results (\figref{p4:fig3}A, B). Thus, increasing fixation depth does not influence self-motion perception during body fixed fixations, but causes self-motion to be perceived as shorter during world-fixed fixations.

@@

In order to quantify how much fixation depth scales the eye movement as a linear self-motion cue, we used a simple linear model (see \nameref{p4:sec:methods}). This model explains the perceived translation distance as a weighted average of a vestibular estimate, equal to the actual translation, and an oculomotor-based estimate. 
We used one weighting parameter, $\alpha$, per fixation depth, $\alpha_{50}$ and $\alpha_{200}$ (see \tabref{p4:tab2} for best-fit values). Using these parameters we predicted the PSEs, i.e. $m_p$ in \eqnref{p4:eq5}, and plotted it against the actually  observed PSEs in \figref{p4:fig4}. The positive correlation ($\rho = 0.xx$, $p < 0.01$) between observed and predicted PSEs shows that our simple model does reasonably well in predicting perceptual performance.

%%%%
